### 2025-05-27 docs/article/Sections/state-of-the-art-web-servers.tex
Section 2 - Web Servers - minor fixes

In traditional thread-per-request architectures, each incoming request is
handled by a dedicated thread. The web server maintains a thread pool from
which a thread is assigned to handle each request. As the load increases, the
number of active threads can grow rapidly, potentially exhausting the pool.
This can lead to performance and scalability issues, as the system may become
bogged down by context switching and thread management
overhead~\cite{kant2000scalable}. In the Java ecosystem, Spring Boot is one of
the most widely used frameworks that follow this model, according to several
reports such as the Stack Overflow 2024 Developer
Survey~\footnote{~\url{https://survey.stackoverflow.co/2024/technology\#1-web-frameworks-and-technologies}}.

@@

The \textit{non-blocking} I/O model employed in low-thread servers is
well-suited for handling large volumes of data asynchronously~\cite{Meijer12}.
This combination of low-thread servers and asynchronous data models has
facilitated the development of highly scalable, responsive, and resilient web
applications capable of managing substantial data loads~\cite{Jin15}. The
prominence of this concept increased with the advent of Node.js in 2009, and
subsequently, various technologies adopted this approach in the Java ecosystem,
including Netty, Akka HTTP, Vert.X, and Spring WebFlux.
% Among these, Spring WebFlux stands out as the most widely used web framework in
% Java, according to surveys like JetBrains' State of Java report (2021) and
% community metrics from platforms like Github and StackOverflow.
The \textit{non-blocking} I/O model in low-thread servers functions optimally
only when HTTP handlers avoid blocking. Therefore, HTML templates need to be
proficient in dealing with the asynchronous APIs provided by data models. While
most legacy web templates struggle with asynchronous models, DSLs for HTML face
no such limitations, leveraging all constructions available in the host
programming language. However, the unexpected intertwining of asynchronous
handlers' completion and HTML builders' execution may potentially lead to
malformed HTML with an unexpected layout, as demonstrated in the subsequent
subsection~\ref{sec:async-support}.

@@

Only Java virtual threads, introduced in
JEP~444~\footnote{\url{https://openjdk.org/jeps/444}}, adopt a similar approach
to Karsten's proposal, using user-mode threads to preserve a synchronous
programming model while still allowing blocking I/O operations without tying up
platform threads. When a virtual thread performs a blocking I/O operation, the
JVM intercepts the call and transparently parks the virtual thread, freeing the
underlying platform thread to perform other tasks. Once the I/O operation
completes, the virtual thread is rescheduled on a platform thread and resumes
execution. This mechanism is enabled by the JVM's integration with non-blocking
system calls at the OS level, allowing developers to write traditional
synchronous code while benefiting from the scalability of non-blocking I/O.

@@

Controllers in Spring MVC typically return Java objects or models that are
processed by view resolvers in conjunction with template engines like JSP,
Thymeleaf, or FreeMarker. These templates generate the full HTML response in
one pass, which is then sent to the client only after all necessary data has
been gathered and rendered. By default, this architecture limits the ability to
perform \textit{Progressive Server-Side Rendering (PSSR)}, as traditional
rendering waits for all data before sending any output. This results in higher
latency for clients and less responsive page loads, particularly for
data-intensive views. However, Spring MVC provides a mechanism for response
streaming via the \texttt{StreamingResponseBody}\footnote{\url{https://docs.spring.io/spring-framework/docs/current/javadoc-api/org/springframework/web/servlet/mvc/method/annotation/StreamingResponseBody.html}} interface, introduced in
Spring 4.2. When a controller method returns a \texttt{StreamingResponseBody},
Spring writes directly to the \texttt{HttpServletResponse} output stream,
allowing the server to send parts of the response incrementally as they become
available. This is particularly useful for:
\begin{itemize}
  \item Sending large responses without buffering the entire output in memory.
  \item Writing dynamic HTML in fragments as data is fetched or computed.
  \item Reducing Time to First Byte (TTFB) and improving perceived performance.
\end{itemize}

@@

Spring WebFlux is a reactive web framework built on a non-blocking,
event-driven architecture that enables efficient handling of a large number of
concurrent connections with minimal resource usage. Unlike traditional
servlet-based stacks that rely on a thread-per-request model, WebFlux operates
on reactive runtimes like Netty, where a small, fixed-size event loop manages
all I/O events.
Incoming HTTP requests are routed to handler functions or annotated controllers,
which return reactive types such as \texttt{Mono<T>} and
\texttt{Flux<T>}—Publisher implementations provided by the Project Reactor
library~\cite{projectreactor}. These implementations adhere to the Reactive
Streams specification~\cite{ReactiveStreams}, enabling interoperability with
other compliant reactive stream libraries.
A \texttt{Mono} represents a single asynchronous value (or
none)~\cite{promise}, while a \texttt{Flux} represents a stream of zero or more values~\cite{rx-observable}. These
abstractions allow the entire request lifecycle—from controller invocation to
response rendering—to be orchestrated as a chain of non-blocking, asynchronous
operations. I/O-bound tasks such as database queries or external API calls do
not occupy threads during execution, enabling the framework to scale gracefully
under high load and maintain responsiveness even in resource-constrained
environments.

Progressive Server-Side Rendering (PSSR) is natively supported in WebFlux via
\texttt{Flux}-based controllers. When a controller returns a
\texttt{Flux<String>} or another stream of HTML fragments, WebFlux can begin
streaming content to the client as soon as the first elements are emitted.
% This allows browsers to incrementally render parts of the response without waiting
% for the entire payload, significantly reducing Time to First Byte (TTFB) and
% improving user-perceived performance. By integrating PSSR directly into the
% reactive processing pipeline, WebFlux facilitates responsive and resilient web
% applications well-suited for modern microservices architectures.

@@

Quarkus is a modern, cloud-native Java framework that supports both imperative
and reactive programming models. It is built on top of Vert.x,
which provides a non-blocking, event-driven runtime similar to Netty. Incoming
HTTP requests are processed asynchronously on a small, fixed-size event loop
thread pool, allowing the framework to handle many concurrent connections
efficiently without blocking threads. 
Quarkus provides support for reactive APIs like Mutiny~\cite{mutiny2021}, a
modern alternative to Spring Reactor~\cite{projectreactor}. Like Reactor,
Mutiny complies with the Reactive Streams specification~\cite{ReactiveStreams}.
For traditional imperative-style request handling, Quarkus provides integration
with JAX-RS, where resource methods can return synchronous types or
asynchronous constructs such as \texttt{CompletionStage}. To enable Progressive
Server-Side Rendering (PSSR) in imperative endpoints, Quarkus leverages the
JAX-RS \texttt{StreamingOutput} interface. By returning a
\texttt{StreamingOutput} instance, developers can write parts of the HTTP
response body incrementally as data becomes available. This approach allows the
server to flush HTML fragments progressively to the client, improving the
perceived responsiveness especially for long-running or large responses.
However, like other streaming mechanisms based on servlet buffers, the initial
data flush may be delayed until internal buffers (commonly around 8KB) are
filled. However, Quarkus allows configuring the buffer size to a smaller value, which
can help mitigate this issue.

\begin{lstlisting}[
  language=java,
  basicstyle=\scriptsize\ttfamily,
  numbers=none,
  caption={Progressive Server-Side Rendering in Quarkus using StreamingOutput},
  label={lst:pssr-quarkus}
]
@GET
@Path("/stream")
@Produces(MediaType.TEXT_HTML)
public StreamingOutput streamHtml() {
    return output -> {
        PrintWriter writer = new PrintWriter(output);
        writer.println("<html><body>");
        writer.flush(); // Flush initial HTML header
        for (int i = 1; i <= 5; i++) {
            writer.println("<p>Chunk " + i + "</p>");
            writer.flush(); // Flush each chunk incrementally
            Thread.sleep(500); // Simulate processing delay
        }
        writer.println("</body></html>");
        writer.flush();
    };
}
\end{lstlisting}

>>>>

In traditional thread-per-request architectures, each incoming request is
handled by a dedicated thread. The web server maintains a thread pool from
which a thread is assigned to handle each request. As the load increases, the
number of active threads can grow rapidly, potentially exhausting the pool.
This can lead to performance and scalability issues, as the system may become
bogged down by context switching and thread management
overhead~\cite{kant2000scalable}. In the Java ecosystem, Spring MVC is one of
the most widely used frameworks that follow this model, according to several
reports such as the Stack Overflow 2024 Developer
Survey~\footnote{~\url{https://survey.stackoverflow.co/2024/technology\#1-web-frameworks-and-technologies}}.

@@

The \textit{non-blocking} I/O model employed in low-thread servers is
well-suited for handling large volumes of data asynchronously~\cite{Meijer12}.
This combination of low-thread servers and asynchronous data models has
facilitated the development of highly scalable, responsive, and resilient web
applications capable of managing substantial data loads~\cite{Jin15}. The
prominence of this concept increased with the advent of Node.js in 2009, and
subsequently, various technologies adopted this approach in the Java ecosystem,
including Netty, Akka HTTP, Vert.X, and Spring WebFlux.
% Among these, Spring WebFlux stands out as the most widely used web framework in
% Java, according to surveys like JetBrains' State of Java report (2021) and
% community metrics from platforms like Github and StackOverflow.
The \textit{non-blocking} I/O model in low-thread servers functions optimally
only when HTTP handlers avoid blocking. Therefore, HTML templates need to be
proficient in dealing with the asynchronous APIs provided by data models. While
most legacy web templates struggle with asynchronous models, DSLs for HTML face
no such limitations, leveraging all constructions available in the host
programming language. However, the unexpected intertwining of asynchronous
handlers' completion and HTML builders' execution may potentially lead to
malformed HTML with an unexpected layout~\cite{wise2024pssr}.

@@

Only Java virtual threads, introduced in
JEP~444\footnote{\url{https://openjdk.org/jeps/444}}, adopt a similar approach
to Karsten's proposal, using user-mode threads to preserve a synchronous
programming model while still allowing blocking I/O operations without tying up
platform threads. When a virtual thread performs a blocking I/O operation, the
JVM intercepts the call and transparently parks the virtual thread, freeing the
underlying platform thread to perform other tasks. Once the I/O operation
completes, the virtual thread is rescheduled on a platform thread and resumes
execution. This mechanism is enabled by the JVM's integration with non-blocking
system calls at the OS level, allowing developers to write traditional
synchronous code while benefiting from the scalability of non-blocking I/O.

@@

Controllers in Spring MVC typically return Java objects or models that are
processed by view resolvers in conjunction with template engines like JSP,
Thymeleaf, or FreeMarker. These templates generate the full HTML response in
one pass, which is then sent to the client only after all necessary data has
been gathered and rendered. By default, this architecture limits the ability to
perform PSSR, as traditional
rendering waits for all data before sending any output. This results in higher
latency for clients and less responsive page loads, particularly for
data-intensive views. However, Spring MVC provides a mechanism for response
streaming via the \texttt{StreamingResponseBody}\footnote{\url{https://docs.spring.io/spring-framework/docs/current/javadoc-api/org/springframework/web/servlet/mvc/method/annotation/StreamingResponseBody.html}} interface, introduced in
Spring 4.2. When a controller method returns a \texttt{StreamingResponseBody},
Spring writes directly to the \texttt{HttpServletResponse} output stream,
allowing the server to send parts of the response incrementally as they become
available. This is particularly useful for:
\begin{itemize}
  \item Sending large responses without buffering the entire output in memory.
  \item Writing dynamic HTML in fragments as data is fetched or computed.
  \item Reducing Time to First Byte (TTFB) and improving perceived performance.
\end{itemize}

@@

Spring WebFlux is a reactive web framework built on a non-blocking,
event-driven architecture that enables efficient handling of a large number of
concurrent connections with minimal resource usage. Unlike traditional
servlet-based stacks that rely on a thread-per-request model, WebFlux operates
on reactive runtimes like Netty, where a small, fixed-size event loop manages
all I/O events.
Incoming HTTP requests are routed to handler functions or annotated controllers,
which return reactive types such as \texttt{Mono<T>} and
\texttt{Flux<T>}—\texttt{Publisher} implementations provided by the Project Reactor
library~\cite{projectreactor}. These implementations adhere to the Reactive
Streams specification~\cite{ReactiveStreams}, enabling interoperability with
other compliant reactive stream libraries.
A \texttt{Mono} represents a single asynchronous value (or
none)~\cite{promise}, while a \texttt{Flux} represents a stream of zero or more values~\cite{rx-observable}. These
abstractions allow the entire request lifecycle—from controller invocation to
response rendering—to be orchestrated as a chain of non-blocking, asynchronous
operations. I/O-bound tasks such as database queries or external API calls do
not occupy threads during execution, enabling the framework to scale gracefully
under high load and maintain responsiveness even in resource-constrained
environments.

PSSR is natively supported in WebFlux via
\texttt{Flux}-based controllers. When a controller returns a
\texttt{Flux<String>} or another stream of HTML fragments, WebFlux can begin
streaming content to the client as soon as the first elements are emitted.
% This allows browsers to incrementally render parts of the response without waiting
% for the entire payload, significantly reducing Time to First Byte (TTFB) and
% improving user-perceived performance. By integrating PSSR directly into the
% reactive processing pipeline, WebFlux facilitates responsive and resilient web
% applications well-suited for modern microservices architectures.

@@

Quarkus is a modern, cloud-native Java framework that supports both imperative
and reactive programming models. It is built on top of Vert.x,
which provides a non-blocking, event-driven runtime similar to Netty. Incoming
HTTP requests are processed asynchronously on a small, fixed-size event loop
thread pool, allowing the framework to handle many concurrent connections
efficiently without blocking threads. 
Quarkus provides support for reactive APIs like Mutiny~\cite{mutiny2021}, a
modern alternative to Spring Reactor~\cite{projectreactor}. Like Reactor,
Mutiny complies with the Reactive Streams specification~\cite{ReactiveStreams}.
For traditional imperative-style request handling, Quarkus provides integration
with JAX-RS, where resource methods can return synchronous types or
asynchronous constructs such as \texttt{CompletionStage}. To enable 
PSSR in imperative endpoints, Quarkus leverages the
JAX-RS \texttt{StreamingOutput} interface. By returning a
\texttt{StreamingOutput} instance, developers can write parts of the HTTP
response body incrementally as data becomes available. This approach allows the
server to flush HTML fragments progressively to the client, improving the
perceived responsiveness especially for long-running or large responses.
However, like other streaming mechanisms based on servlet buffers, the initial
data flush may be delayed until internal buffers (commonly around 8KB) are
filled. However, Quarkus allows configuring the buffer size to a smaller value, which
can help mitigate this issue.

\begin{lstlisting}[
  language=java,
  basicstyle=\scriptsize\ttfamily,
  numbers=none,
  caption={Progressive Server-Side Rendering in Quarkus using StreamingOutput},
  label={lst:pssr-quarkus}
]
@GET
@Path("/stream")
@Produces(MediaType.TEXT_HTML)
public StreamingOutput streamHtml() {
    return output -> {
        PrintWriter writer = new PrintWriter(output);
        for (int i = 1; i <= 5; i++) {
            writer.println("<p>Chunk " + i + "</p>");
            writer.flush(); // Flush each chunk incrementally
            Thread.sleep(500); // Simulate processing delay
        }
        writer.flush();
    };
}
\end{lstlisting}
